# Human DLPFC HVGs vs SVGs from `nnSVG` 

## Required Packages
```{r include=FALSE}
library(scater)
library(scran)
library(nnSVG)
library(ggspavis)
library(dplyr)
library(tibble)
library(tidyr)
```


## Data

```{r}
# Load from datasets folder
spe <- humanDLPFC
spe
```

## QC

```{r}
# library(scater)

# subset to keep only spots over tissue
spe <- spe[, colData(spe)$in_tissue == 1]

# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)

# calculate per-spot QC metrics and store in colData
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))

# select QC threshold for library size
qc_lib_size <- colData(spe)$sum < 600
colData(spe)$qc_lib_size <- qc_lib_size

# select QC threshold for number of expressed genes
qc_detected <- colData(spe)$detected < 400
colData(spe)$qc_detected <- qc_detected

# select QC threshold for mitochondrial read proportion
qc_mito <- colData(spe)$subsets_mito_percent > 28
colData(spe)$qc_mito <- qc_mito

# select QC threshold for number of cells per spot
qc_cell_count <- colData(spe)$cell_count > 10
colData(spe)$qc_cell_count <- qc_cell_count

# combined set of discarded spots
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
# store in object
colData(spe)$discard <- discard

# remove combined set of low-quality spots
spe <- spe[, !colData(spe)$discard]

dim(spe)
```

```{r}
## View spe after QC
spe
```

## Normalization

```{r}
# library(scran)

# calculate library size factors
spe <- computeLibraryFactors(spe)

hist(sizeFactors(spe), breaks = 20)
```

```{r}
spe <-  logNormCounts(spe)
```


## Feature selection 


```{r}
# remove mitochondrial genes
spe <- spe[!is_mito, ]
```


### SVGs (`nnSVG`)

#### SUBSAMPLE FOR FASTER RUN TIME - USE FULL SAMPLE ON `SCC`!!!!! & USE DEFAULT FILTERING FOR FULL ANALYSIS!!!!
```{r}
# library(nnSVG)

# subsample spots for faster runtime in this example
# note: skip this step in full analysis
n <- 100
set.seed(123)
ix <- sample(seq_len(n), n)
spe_nnSVG <- spe[, ix]

# filter low-expressed and mitochondrial genes
# using stringent filtering for faster runtime in this example
# note: use default filtering in full analysis
spe_nnSVG <- filter_genes(
  spe_nnSVG, filter_genes_ncounts = 10, filter_genes_pcspots = 3
)
```


```{r}
# library(nnSVG)

# re-calculate logcounts after filtering
spe_nnSVG <- logNormCounts(spe_nnSVG)
```


```{r}
# run nnSVG
set.seed(123)
spe_nnSVG <- nnSVG(spe_nnSVG)
```

  - started around 12:45ish?, ended 12:48
  
```{r}
# show results for top n SVGs
rowData(spe_nnSVG)[order(rowData(spe_nnSVG)$rank)[1:6], ]
```
  
```{r}
nrow(rowData(spe_nnSVG)[rowData(spe_nnSVG)$padj<0.05,])
```


### HVGs

```{r}
spe_HVG <- logNormCounts(spe_nnSVG)
```

```{r}
#mlibrary(scran)

# fit mean-variance relationship
dec <- modelGeneVar(spe_HVG)

# select top HVGs
dim(spe_HVG)
top_hvgs <- getTopHVGs(dec, prop = 0.1)
length(top_hvgs)
```

```{r}
dec <- dec[order(dec$p.value),]
n <- nrow(dec)
dec$rank <- 1:n
head(dec)
```


```{r}
spe_topHVGs <- spe_HVG[rowData(spe_HVG)$gene_id %in% top_hvgs,]
```

```{r}
spe_topHVGs
```

### How many genes are both SVGs and HVGs

```{r}
intersect(rowData(spe_nnSVG)$gene_id, rowData(spe_topHVGs)$gene_id)
```


### Correlation between Ranks

```{r}
hvg_ranks <- as.data.frame(dec) %>% rownames_to_column(var = "gene_id") %>% select(gene_id, rank)

svg_ranks <- as.data.frame(rowData(spe_nnSVG)) %>% select(gene_id, rank, gene_name)

rank_df <- merge(hvg_ranks, svg_ranks, by = "gene_id", suffixes = c("_hvgs", "_svgs"))
head(rank_df)
```


```{r}
# plot
ggplot(rank_df, aes(x=rank_hvgs, y=rank_svgs)) +
  geom_point() +
  smplot2::sm_statCorr(color = 'dodgerblue', corr_method = 'spearman')
```


### Proportion overlap of top _n_ ranked highly vs spatially VGs

  - 10, 20, 30, 40, 50


```{r}
#res_nnSVG <- as.data.frame(rowData(spe_nnSVG)) %>% arrange(rank) %>% 

res_list <- list(
  humanDLPFC_nnSVG = rowData(spe_nnSVG), 
  humanDLPFC_HVGs = dec
)

# add method names to all columns except gene IDs and gene names
colnames(res_list[["humanDLPFC_nnSVG"]])[-(1:2)] <- paste0(colnames(res_list[["humanDLPFC_nnSVG"]]), "_nnSVG")[-(1:2)]
colnames(res_list[["humanDLPFC_HVGs"]])[-(1:2)] <- paste0(colnames(res_list[["humanDLPFC_HVGs"]]), "_HVGs")[-(1:2)]

```



```{r}
overlaps <- c(10, 20, 30, 40, 50)
```


`calc_overlaps`
```{r}
calc_overlaps <- function(method1, method2) {
  
  res_method1 <- res_list[[method1]]
  res_method2 <- res_list[[method2]]
  
  # remove method names from column names
  colnames(res_method1)[-(1:2)] <- gsub("_.*$", "", colnames(res_method1)[-(1:2)])
  colnames(res_method2)[-(1:2)] <- gsub("_.*$", "", colnames(res_method2)[-(1:2)])
  
  top_method2 <- rep(NA, length(overlaps))
  
  for (k in seq_along(overlaps)) {
    # select top gene IDs from method 1
    genes_k <- rownames(filter(as.data.frame(res_method1), 
                               rank <= overlaps[k]))
    # calculate overlaps
    top_method2[k] <- nrow(filter(as.data.frame(na.omit(res_method2[genes_k, ])), 
                                  rank <= overlaps[k]))
  }
  
  # calculate proportions
  return(top_method2 / overlaps)
}
```

```{r}
df_overlaps <- data.frame(
  top_n = overlaps, 
  dataset = "humanDLPFC", 
  nnSVG = calc_overlaps("humanDLPFC_HVGs", "humanDLPFC_nnSVG")
)

df_overlaps <- 
  pivot_longer(df_overlaps, cols = "nnSVG", 
               names_to = "method", values_to = "proportion")

```

```{r}
ggplot(as.data.frame(df_overlaps), 
       aes(x = top_n, y = proportion)) + 
  geom_line(lwd = 0.75) + 
  geom_point(size = 2) + 
  scale_x_continuous(breaks = overlaps) + 
  ylim(c(0, 1)) + 
  xlab("top n genes") + 
  ylab("proportion overlap") + 
  ggtitle("Overlap SVGs and HVGs: human DLPFC") + 
  theme_bw() + 
  theme(panel.grid.minor = element_blank())
```



## Dimensionality Reduction


### SVGs

```{r}
# library(scater)

# compute PCA
set.seed(123)
spe_nnSVG <- runPCA(spe_nnSVG, ncomponents=2)

reducedDimNames(spe_nnSVG)
```


### HVGs

```{r}
# library(scater)

# compute PCA
set.seed(123)
spe_topHVGs <- runPCA(spe_HVG, ncomponents = 2, subset_row = top_hvgs)

reducedDimNames(spe_topHVGs)
```


```{r}
# library(ggspavis)
# plot top 2 PCA dimensions
plotDimRed(spe_topHVGs, plot_type = "PCA")
```


## Clustering

### SVGs

```{r}
# graph-based clustering
set.seed(123)
k <- 10
g <- buildSNNGraph(spe_nnSVG, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
table(clus)
```

```{r}
# store cluster labels in column 'label' in colData
colLabels(spe_nnSVG) <- factor(clus)
```


```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe_nnSVG, annotate = "label", 
          pal = "libd_layer_colors")
```


### HVGs

```{r}
# graph-based clustering
set.seed(123)
k <- 10
g <- buildSNNGraph(spe_topHVGs, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
table(clus)
```

```{r}
# store cluster labels in column 'label' in colData
colLabels(spe_topHVGs) <- factor(clus)
```


```{r}
# plot clusters in spatial x-y coordinates
plotSpots(spe_topHVGs, annotate = "label", 
          pal = "libd_layer_colors")
```


### Clustering performance metrics


